Bayesian Detection of Bias in Peremptory Challenges Using Historical Strike Data

### 2023-02-17 analysis/memo_TAS_reviewers.Rmd
revised TAS memo - more on hyperprior for alpha etc.; updated references, minor edit to paper

---
output: 
  pdf_document:
    citation_package: natbib
    keep_tex: false
    latex_engine: pdflatex
    template: "svm-latex-memo.tex"
    number_sections: true
fontfamily: mathpazo
fontsize: 11pt
geometry: margin=1in
header-includes:
   - \linespread{1.05}

from: Authors
to: The American Statistican
subject: Reply to Reviewer Comments
date: "`r format(Sys.Date(), '%B %d, %Y')`"
memorandum: true
graphics: false
width: .3
---

@@

1. As our revised Introduction now makes clear (p. 4-5 and Fig. 1), neither our statistical approach nor any other can alone detect _Batson_ violations, because _Batson_ procedure permits any type of admissible evidence of illegal strike bias -- not just the attorney's past strike data -- and requires the trial judge to weigh all such evidence to decide whether illegal strike bias exists. For example, _Batson_ procedure requires the striking attorney at a certain stage to justify the challenged strikes based on (non-discriminatory) reasons. If the attorney fails to produce such reasons, the trial judge must find a _Batson_ violation. If the attorney offers such reasons, the trial judge must assess evidence as to the credibility of those reasons. Our statistical approach cannot adjust the posterior distribution of the bias parameter to account for the credibility of those reasons or other types of evidence (other than the attorney's pattern of strikes) of illegal strike bias. Rather, our approach helps attorneys evaluate the inference of illegal bias in the present case based on historical strike data but _before_ accounting for other evidence relevant to illegal strike bias. The phrase "bias detection" in the paper means only that the credible interval for the bias parameter excludes zero, given the historical strike data, the model, and choice of $\alpha$ value of the power prior. For that "detection" task, our approach lets one assign at most equal weight to past strikes and their strikes in the present case ($\alpha = 1$). In turn, attorneys can use the resulting bias estimate, combined with other evidence, to (1) decide whether to argue _Batson_, (2) to prove the _Batson_ prima facie case, and (3) to persuade the trial judge to find  a _Batson_ violation (i.e., illegal strike bias is more likely than not, given all the evidence).

@@

4. Absent some very strong assumptions, there is no generally justifiable hyper prior for $\alpha$ in this context of strike bias and _Batson_ challenges. This is because the proper weight to place on historical data depends on what causes incompatibility between the historical data and the current data. The plausible reasons for non-random missingness and such confounding, however, likely vary considerably depending on strike procedure, courtroom practices, prosecutor charging practices, among other things. As we now note in the Discussion section of the paper, measuring incompatibility and choosing $\alpha$ in the face of incompatibility is an active area of research with no clear solution.

@@

We have three replies to this concern.

First, the law of _Batson_ and parallel State law has consistently permitted judges to infer illegal strike bias based in part on an attorney's use of strikes in past cases. In reviewing the kinds of evidence that can prove a _Batson_ violation by a prosecutor, the U.S. Supreme Court included, as examples of types of evidence, the "relevant history of the State’s peremptory strikes in past cases." _Flowers v. Mississippi_, 139 S. Ct. 2228, [PIN] (2019). State law variants on _Batson_ are also explicit about this. For example, Washington State law provides that in deciding whether "an objective observer could view race or ethnicity as a factor in the use of the peremptory challenge," the circumstances the court "should" consider include "whether the party has used peremptory challenges disproportionately against a given race or ethnicity, in the present case _or in past cases_." Wash. General Rule 37(g)(v) (emphasis added). California law similarly identifies, as a circumstance a court may consider, whether the striking party "has used peremptory challenges disproportionately against a given race . . . in the present case _or in past cases_." Calif. Code of Civil Procedure 231.7(d)(3)(G) (emphasis added).

@@

[editme]In any case, other authors analysing _Batson_ challallenge data have suggest other, more implicit forms of weighting cases based on trial-incompatibility. As we now mention in the paper [xx], @Gastwirth2014 suggests that trials with 'similar characteristics' be selected before conducting Cochran-Mantel-Haenszel tests. This is tantamount to setting the weight of incompatible cases to zero and compatible cases to 1. In such an instance, such weighing should be explicit and justified.

@@

Part of the concern here may turn on an error about the standard of proof, as evinced by the use of the word "guilty" in this comment.  _Batson_ violations must be proven under the "preponderance of the evidence" standard (more likely than not), _not_ the more stringent "reasonable doubt" standard that applies when a prosecutor aims to prove a defendant "guilty" of committing the criminal offense charged.

@@

We revised to remove such confusion. On the choice of credible intervals, see Item 1(5) above. The revision does not necessarily fix the type I error, because this coverage probability refers to the true bias parameter (that is fixed in our simulations), whereas we are interested in "bias detection", i.e. whether the credible interval includes 0 or not. The paper also has a sensitivity analysis for the selection of the nominal credible interval level.

@@

This comment is not clear to us. The real data analyses are based on real historical strike data from real criminal cases in the federal district court in Connecticut. If they wish, users of the app can run the model excluding any historical data (as this reviewer suggested) by not selecting the name of a prosecutor or defense attorney. Like our model, the app uses the $N(0,2)$ prior for the bias parameter. Similarly, the app user can alter the weight of the historical strike data (if selected) by adjusting $\alpha$ (the power prior parameter). Thus, the app user can: (1) choose not to use historical data, (2) use historical data with equal weight as the current trial (alpha = 1), or (3) down-weight the historical trial data (alpha = 0.5 or alpha = 0.2). See also Item 1(6).

@@

We respectfully disagree with the analogy to "p-hacking". The phrase "p-hacking" typically refers to the practice of using different statistical models on the same data in search of one or more associations that qualify as "statistically significant" (p < .05), and then reporting only the analysis with the model(s) that produce those results. In contrast, in our approach, the statistical model has to qualify as a valid model of the strike procedure of the court of interest. That strike procedure can and must be known and modeled before any strike data is collected or analyzed. More generally, our approach thwarts the strategic selection of $\alpha$ values by making the choice of $\alpha$ transparent. See Item 1(4) and section 3.2 of the revised Discussion.

@@

On confounders, we expressly highlight that variation in trial-level characteristics (e.g. defendant race, charge severity) in the historical data can generate incompatibility, which in turn can affect bias detection. However, with respect to juror-level characteristics, the comment does not apply. Suppose a prosecutor strikes three Black prospective jurors on the belief that Black people are generally less educated and therefore more likely to favor defendants. This is illegal under _Batson_, regardless of the actual education levels of the three jurors struck.

On unintended consequences, we found this comment unclear. The paper aims to show the validity of our Bayesian approach for incorporating historical strike data for estimating strike bias in the context of _Batson_ challenges. We do not aim to identify any and all consequences of a full-scale implementation of this approach, intended or unintended.

@@

We agree that, given high incompatibility, (1) the bias "detection" rate can be as high as 100% when current bias is low ($b_curr$); and (2) bias detection can be low where current bias is high but bias in historical trials is low. The tradeoff is that, given compatibility, historical strike data increases power that is otherwise low due the limited number of strikes per trial. The advantage of using the power prior: We can adjust the weight of the historical strike data (and therefore type I error) by adjusting $\alpha$.

@@

We appreciate this suggestion but we are worried that this would unduly increase the paper's length. Because of this concern, we have already moved other parts of the paper to the (online) Appendix.

@@

We assumed no strike bias for the defense attorney only to simplify exposition of the simulation results. Our model of the strike procedure (Equation 1) assumes that the probability of prosecutor strike bias is independent of the probability of defense attorney strike bias. And that model explicitly accounts for the racial composition of the pool of prospective jurors at each strike opportunity.

@@

In any case, we re-ran our simulation and found that, as expected, changing the level of defense attorney bias does not alter the bias detection rates for prosecutors (Table 1).

@@

No, this is not correct. We have revised the paper to avoid such confusion. See Item 1(5) above. We agree that the choice of credible interval should be consistent within different stages of the _Batson_ challenge and across judges deciding _Batson_ challenges within the same jurisdiction. [NOTE: Simulations use 95% credible interval, while app uses 80% credible intervals.]

@@

On the choice of $\alpha$, see Items 1(2) and 1(3) above and section 3.1 of the revised paper. At most, the historical strike data and the current strike data have _equal_ weight (when $\alpha$ = 1).

@@

>>>>

---
output: 
  pdf_document:
    citation_package: biblatex
    keep_tex: false
    latex_engine: pdflatex
    template: svm-latex-memo.tex
    number_sections: true
fontfamily: mathpazo
fontsize: 11pt
geometry: margin=1in
header-includes:
   - \linespread{1.05}

from: Authors
to: The American Statistican
subject: Reply to Reviewer Comments
date: "`r format(Sys.Date(), '%B %d, %Y')`"
memorandum: true
graphics: false
width: .3
bibliography: references.bib
---

@@

1. As our revised Introduction now makes clear (p. 4-5 and Fig. 1), neither our statistical approach (nor any other) can alone detect _Batson_ violations, because _Batson_ procedure permits any type of admissible evidence of illegal strike bias -- not just the attorney's past strike data -- and requires the trial judge to weigh all such evidence to decide whether illegal strike bias exists.

    a. For example, _Batson_ procedure requires the striking attorney at a certain stage to justify the challenged strikes based on (non-discriminatory) reasons. If the attorney fails to produce such reasons, the trial judge must find a _Batson_ violation. If the attorney offers such reasons, the trial judge must assess evidence as to the credibility of those reasons. Our statistical approach cannot adjust the posterior distribution of the bias parameter to account for the credibility of those reasons or other types of evidence (other than the attorney's pattern of strikes) of illegal strike bias. Rather, our approach helps attorneys evaluate the inference of illegal bias in the present case based on historical strike data but _before_ accounting for other evidence relevant to illegal strike bias.

    b. The phrase "bias detection" in the paper means only that the credible interval for the bias parameter excludes zero, given the historical strike data, the model, and choice of $\alpha$ value of the power prior. For that "detection" task, our approach lets one assign at most equal weight to past strikes and their strikes in the present case ($\alpha = 1$). In turn, attorneys can use the resulting bias estimate, combined with other evidence, to (1) decide whether to argue _Batson_, (2) to prove the _Batson_ prima facie case, and (3) to persuade the trial judge to find  a _Batson_ violation (i.e., illegal strike bias is more likely than not, given all the evidence).

@@

4. Absent some very strong assumptions, there is no generally justifiable hyper prior for $\alpha$ in this context of strike bias and _Batson_ challenges. This is because the proper weight to place on historical data depends on what causes incompatibility between the historical data and the current data. The plausible reasons for non-random missingness and such confounding, however, likely vary considerably depending on strike procedure, courtroom practices, prosecutor charging practices, among other things. As we now note in the Discussion section of the paper, the literature on the power prior discusses information criteria for guide values of $\alpha$ as well as treating $\alpha$ as random and assigning a prior to it. However, that literature is unclear as to what implies for _Batson_ and peremptory strikes, for at least three reasons.  First, that literature has mostly focused on clinical trials and models whose properties are well known (e.g., the linear model), as distinct from the models of strike procedure discussed here.  Second, our approach does not attempt to maximize coverage in the traditional sense, but aims only to "detect" bias by determining whether zero falls outside the chosen credible interval for the posterior distribution of the bias parameter. Third, whereas some papers treat incompatibility as itself a parameter [e.g., @Ollier2020], the precise approaches rely on strong assumptions that do not obviously make sense in our context.  Accordingly, we believe the most conservative choice (for now) is to keep $\alpha$ a fixed value and check for sensitivity of results to the choice of $\alpha$ value. Proceeding otherwise would require the work of a completely new paper.

@@

We have four replies to this concern.

First, the law of _Batson_ and parallel State law has consistently permitted judges to infer illegal strike bias based in part on an attorney's use of strikes in past cases. In reviewing the kinds of evidence that can prove a _Batson_ violation by a prosecutor, the U.S. Supreme Court included, as examples of types of evidence, the "relevant history of the State’s peremptory strikes in past cases." _Flowers v. Mississippi_, 139 S. Ct. 2228, 2243 (2019). State law variants on _Batson_ are also explicit about this. For example, Washington State law provides that in deciding whether "an objective observer could view race or ethnicity as a factor in the use of the peremptory challenge," the circumstances the court "should" consider include "whether the party has used peremptory challenges disproportionately against a given race or ethnicity, in the present case _or in past cases_." Wash. General Rule 37(g)(v) (emphasis added). California law similarly identifies, as a circumstance a court may consider, whether the striking party "has used peremptory challenges disproportionately against a given race . . . in the present case _or in past cases_." Calif. Code of Civil Procedure 231.7(d)(3)(G) (emphasis added).

@@

Fourth, the comment is overinclusive. Most work on the use of historical strike data (a) presupposes that _Batson_ law permits prior strikes as relevant evidence of bias in the current trial; presupposes that any statistical approach therein cannot automatically determine _Batson_ violations alone; and (c) adjusts in some way for incompatibility between historical strike data and strikes from the current trial. As an example, we now point in the paper's Introduction to @Gastwirth2014. Our approach has a distinctive benefit of making incompatibility assumptions transparent via choosing a fixed value for the power prior's $\alpha$.

@@

The use of the word "guilty" in this comment implies that part of the concern here may turn on an error about the applicable standard of proof.  _Batson_ violations must be proven under the "preponderance of the evidence" standard (more likely than not), _not_ the more demanding  "reasonable doubt" standard that applies when a prosecutor aims to prove a defendant "guilty" of committing the elements of a criminal offense.

@@

We revised to remove such confusion. This includes changing the software app to use 95% credible intervals. On the choice of credible intervals, see Item 1(5) above. The revision does not necessarily fix the type I error, because this coverage probability refers to the true bias parameter (that is fixed in our simulations), whereas we are interested in "bias detection", i.e. whether the credible interval includes 0 or not. The paper also has a sensitivity analysis for the selection of the nominal credible interval level.

@@

This comment is not clear to us. The real data analyses are based on real historical strike data from real criminal cases in the federal district court in Connecticut. On how we collected this data, see the paper's Appendix. If they wish, users of the app can run the model excluding any historical data (as this reviewer suggested) by not selecting the name of a prosecutor or defense attorney. Like our model, the app uses the $N(0,2)$ prior for the bias parameter. Similarly, the app user can alter the weight of the historical strike data (if selected) by adjusting $\alpha$ (the power prior parameter). Thus, the app user can: (1) choose not to use historical data, (2) use historical data with equal weight as the current trial (alpha = 1), or (3) down-weight the historical trial data (alpha = 0.5 or alpha = 0.2). See also Item 1(6).

@@

We respectfully disagree with the analogy to "p-hacking". The phrase "p-hacking" typically refers to the practice of using different statistical models on the same data in search of one or more associations that qualify as "statistically significant" (p < .05), and then reporting only the analysis with the model(s) that produce those results. In contrast, in our approach, the statistical model has to qualify as a valid model of the strike procedure of the court of interest. That strike procedure must be known and modeled before any strike data is analyzed. More generally, our approach thwarts the strategic selection of $\alpha$ values by making the choice of $\alpha$ transparent. See Item 1(4) and section 3.2 of the revised Discussion.

@@

On confounders, we expressly highlight (in the Introduction) that variation in trial-level characteristics (e.g. defendant race, charge severity) in the historical data can generate incompatibility, which in turn can affect bias detection. However, with respect to juror-level characteristics, the comment does not apply. Suppose a prosecutor strikes three Black prospective jurors on the belief that Black people are generally less educated and therefore more likely to favor defendants. This is illegal under _Batson_, regardless of the actual education levels of the three jurors struck.

On unintended consequences, we found this comment unclear. The paper aims to show the validity of our Bayesian approach for incorporating historical strike data for estimating strike bias in the context of _Batson_ challenges. We do not aim to identify any and all consequences of a full-scale implementation of this approach, intended or unintended. Such implementation entails a series of operational choices that are well beyond the paper's scope.

@@

We agree that, given high incompatibility, (1) the bias "detection" rate can be as high as 100% when current bias is low ($b_curr$); and (2) bias detection can be low where current bias is high but bias in historical trials is low. The tradeoff is that, given high compatibility, historical strike data increases power that is otherwise low due the limited number of strikes per trial. The advantage of using the power prior: We can adjust the weight of the historical strike data (and therefore type I error) by adjusting $\alpha$.

@@

We appreciate this suggestion but we are worried that this would unduly increase the paper's length, particular given that absence of a closed form solution for the posterior distribution even in a toy example. Because of this concern, we have already moved other parts of the paper to the (online) Appendix. We intend the reader to gain understanding by following along using our prototype software application.

@@

We assumed no strike bias for the defense attorney only to simplify exposition of the simulation results. Our model of the strike procedure (Equation 1) assumes that the probability of prosecutor strike bias is independent of the probability of defense attorney strike bias. And that model explicitly accounts for the racial composition of the pool of prospective jurors at each strike opportunity. In any case, we re-ran our simulation and found that, as expected, changing the level of defense attorney bias does not alter the bias detection rates for prosecutors (Table 1).

@@

@@

No, this is not correct. We have revised the paper and the software application to avoid such confusion. See Item 1(5) above. We agree that the choice of credible interval should be consistent within different stages of the _Batson_ challenge and across judges deciding _Batson_ challenges within the same jurisdiction.

@@

On the choice of $\alpha$, see Items 1(2)-1(4) above and section 3.1 of the revised paper. At most, the historical strike data and the current strike data have _equal_ weight (when $\alpha$ = 1).

@@

# References
